Explanation-based learning of action models

### 2019-03-28 doc/lecture-notes-EBL/ebl.tex
Final updates

\section{Learning \strips\ action models with classical planning}
Our approach to address a learning task $\Lambda=\tup{\mathcal{M},{\mathcal O}}$ is to compile it into a classical planning problem $P_{\Lambda}$. The intuition behind the compilation is that when $P_{\Lambda}$ is solved, the solution plan $\pi_\Lambda$ is a sequence of actions that build the output model $\mathcal{M'}$ and verify that $\mathcal{M'}$ is consistent with the observation ${\mathcal O}$.

A solution plan $\pi_\Lambda$ comprise then two differentiated blocks of actions: a plan prefix with a set of actions each defining the \textbf{insertion} of a fluent as a {\em precondition} or a {\em effect} of an action model; and a plan postfix with a set of actions that determine the \textbf{application} of the learned modes while successively \textbf{validating} the effects of the action application in every partial state in ${\mathcal O}$. Roughly speaking, in the \emph{blocksworld}, the format of the first set of actions of $\pi_\Lambda$ looks like {\tt{\small (insert\_pre\_stack\_holding\_v1)}, {\tt\small (insert\_eff\_stack\_clear\_v1)}, {\tt\small (insert\_eff\_stack\_holding\_v1)}}\ldots where the first effect denotes a positive effect and the second one a negative effect to be inserted in $name(\xi)=${\tt{\small stack}}; and the format of the second set of actions of $\pi_\Lambda$ is like {\tt{\small (apply\_unstack blockB blockA),(apply\_putdown blockB)}} and {\tt{\small (validate\_1)}, {\tt\small (validate\_2)}}, where the last two actions denote the points at which the states generated through the action application must be validated with the observed states in ${\mathcal O}$.

\subsection{Compilation}
Given a learning task $\Lambda=\tup{\mathcal{M},{\mathcal O}}$ the compilation outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$ such that:
\begin{itemize}
\item $F_{\Lambda}$ extends the set of fluents $F$ (obtained from $s_0^o$) with the model fluents to represent the preconditions and effects of each $\xi\in\mathcal{M}$ as well as some other fluents to keep track of the validation of ${\mathcal O}$. Specifically, $F_{\Lambda}$ contains also:
\begin{itemize}
\item Fluents $pre_{p,\xi}$ and $eff_{p,\xi}$, defined as explained in section~\ref{sec:strips-space}.
\item A set of fluents $\{test_j\}_{0\leq j\leq m}$, to point at the state observation $s_j^o\in {\mathcal O}$ where the action model is validated. In the example of Figure~\ref{fig:observation} two tests are required to validate the programmed action model, one corresponding to the initial state and the second one corresponding to the final state.
\item A fluent, $mode_{prog}$, to indicate whether action models are being programmed or validated and a fluent {\em invalid} to indicate that the programmed action model is inconsistent with the input observation.
\end{itemize}

@@

\item $A_{\Lambda}$ includes three types of actions that give rise to the actions of $\pi_\Lambda$.
\begin{enumerate}
\item Actions for {\em inserting} a precondition or effect into $\xi \in \mathcal{M}$ following the syntactic constraints of \strips\ models. These actions will form the prefix of the solution plan $\pi_\Lambda$. Among the \emph{inserting} actions, we find:
\begin{itemize}
\item Actions for inserting a {\em precondition} $p\in{\mathcal I}_{\xi,\Psi}$ into $\xi$.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertPre_{p,\xi}})=&\{\neg pre_{p,\xi}, mode_{prog}\},\\
\cond(\mathsf{insertPre_{p,\xi}})=&\{\emptyset\}\rhd\{pre_{p,\xi}\}.
\end{align*}
\end{small}

\item Actions for inserting a {\em effect} $p\in{\mathcal I}_{\xi,\Psi}$ into $\xi$.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertEff_{p,\xi}})=&\{\neg eff_{p,\xi}, mode_{prog}\},\\
\cond(\mathsf{insertEff_{p,\xi}})=&\{\emptyset\}\rhd\{eff_{p,\xi}\}
\end{align*}
\end{small}
\end{itemize}

@@

\item Actions for {\em applying} the action models $\xi\in\mathcal{M}$ built by the insert actions and bound to objects $\omega\subseteq\Omega^{ar(\xi)}$. These actions will be part of the postfix of the plan $\pi_\Lambda$ and they determine the application of the learned action models according to the values of the model fluents in the current state configuration. Since action headers are known, the variables $pars(\xi)$ are bounded to the objects in $\omega$ that appear in the same position.

>>>>

\section{Learning \strips\ action models with classical planning}
Our approach to address a learning task $\Lambda=\tup{\mathcal{M},{\mathcal O}}$ is to compile it into a classical planning problem $P_{\Lambda}$. The intuition behind the compilation is that when $P_{\Lambda}$ is solved, the solution plan $\pi_\Lambda$ is a sequence of actions that build the output model $\mathcal{M'}$ and verify that $\mathcal{M'}$ explains the observation ${\mathcal O}$.

A solution plan $\pi_\Lambda$ includes then two differentiated blocks of actions: a plan prefix with a set of actions, each defining the \textbf{insertion} of a fluent as a precondition or an effect of an action model; and a plan postfix with a set of actions that determine the \textbf{application} of the learned modes while successively \textbf{validating} the effects of the action application in every partial state of ${\mathcal O}$. Roughly speaking, in the \emph{blocksworld}, the format of the first block of actions of $\pi_\Lambda$ looks like {\tt{\small (insert\_pre\_stack\_holding\_v1)}, {\tt\small (insert\_eff\_stack\_clear\_v1)}, {\tt\small (insert\_eff\_stack\_holding\_v1)}}\ldots where the first effect denotes a positive effect and the second one a negative fluent to be inserted in $name(\xi)=${\tt{\small stack}}; and the format of the second block of actions of $\pi_\Lambda$ is like {\tt{\small (apply\_unstack blockB blockA),(apply\_putdown blockB)}} and {\tt{\small (validate\_1)}, {\tt\small (validate\_2)}}, where the last two actions denote the points at which the states generated through the action application must be validated with the observed states in ${\mathcal O}$.

\subsection{Compilation}
Given a learning task $\Lambda=\tup{\mathcal{M},{\mathcal O}}$ the compilation outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$ such that:
\begin{itemize}
\item $F_{\Lambda}$ extends the set of fluents $F$ (obtained from $s_0^o$) with the model fluents that are used to represent the preconditions and effects of each $\xi\in\mathcal{M}$ as well as some other fluents to keep track of the validation of ${\mathcal O}$. Specifically, $F_{\Lambda}$ contains also:
\begin{itemize}
\item Fluents $pre_{p,\xi}$ and $eff_{p,\xi}$, defined as explained in section~\ref{sec:strips-space}.
\item A set of fluents $\{test_j\}_{0\leq j\leq m}$, to point at the state observation $s_j^o\in {\mathcal O}$ where the action model is validated. In the example of Figure~\ref{fig:observation} two tests are required to validate the programmed action model, one corresponding to the initial state and the second one corresponding to the final state.
\item A fluent, $mode_{prog}$, to indicate whether action models are being programmed or validated and a fluent {\em invalid} to indicate that the programmed action model is inconsistent with the input observation.
\end{itemize}

@@

\item $A_{\Lambda}$ includes three types of actions that give rise to the actions of $\pi_\Lambda$.
\begin{enumerate}

\item Actions for {\em inserting} a precondition or effect into $\xi \in \mathcal{M}$ following the syntactic constraints of \strips\ models. These actions will form the prefix of the solution plan $\pi_\Lambda$. Among the \emph{inserting} actions, we find:

\begin{itemize}
\item Actions for inserting a {\em precondition} $p\in{\mathcal I}_{\xi,\Psi}$ into $\xi$.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertPre_{p,\xi}})=&\{\neg pre_{p,\xi}, mode_{prog}\},\\
\cond(\mathsf{insertPre_{p,\xi}})=&\{\emptyset\}\rhd\{pre_{p,\xi}\}.
\end{align*}
\end{small}

\item Actions for inserting an {\em effect} $p\in{\mathcal I}_{\xi,\Psi}$ into $\xi$.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertEff_{p,\xi}})=&\{\neg eff_{p,\xi}, mode_{prog}\},\\
\cond(\mathsf{insertEff_{p,\xi}})=&\{\emptyset\}\rhd\{eff_{p,\xi}\}
\end{align*}
\end{small}
\end{itemize}

@@

\item Actions for {\em applying} the action models $\xi\in\mathcal{M}$ built by the insert actions and bounded to objects $\omega\subseteq\Omega^{ar(\xi)}$. These actions will be part of the postfix of the plan $\pi_\Lambda$ and they determine the application of the learned action models according to the values of the model fluents in the current state configuration. Since action headers are known, the variables $pars(\xi)$ are bounded to the objects in $\omega$ that appear in the same position.
